# Load necessary libraries
```{r}
library(tidyverse)
library(patchwork)
```

# Data Import
```{r}
df <- read.csv("./results/model_evaluation_results_allData.csv")
head(df)
```

# Descriptive Statistics

## Setting variables

```{r}
crops <- c("wheat", "sorghum")
model_name <- c("DeepLabV3Plus", "U-NET", "SegNet", "SegFormer", "MaskFormer")
metrics <- c("IoU", "Dice", "Precision", "Recall", "F1")
```

## Data summary
```{r}
summary(df)
```

## When removing outliers
```{r}
remove_outliers <- function(x, na.rm = TRUE, ...) {
    qnt <- quantile(x, probs = c(0.25, 0.75), na.rm = na.rm, ...)
    H <- 1.5 * IQR(x, na.rm = na.rm)
    y <- x
    y[x < (qnt[1] - H)] <- NA
    y[x > (qnt[2] + H)] <- NA
    return(y)
}

df_no_outliers <- df %>%
    mutate(across(all_of(metrics), ~remove_outliers(.x))) %>%
    select(crop, modelname, all_of(metrics), genotype, collectiondate) %>%
    drop_na()

summary(df_no_outliers)

table(df_no_outliers$crop, df_no_outliers$modelname)
```

## Getting Mean, Max, Min, var and cv.
```{r}
# Compute summary statistics by crop and model
summary_stats <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    group_by(crop, modelname) %>%
    summarise(across(all_of(metrics),
                                     list(mean = ~mean(.x, na.rm = TRUE),
                                          max  = ~max(.x, na.rm = TRUE),
                                          min  = ~min(.x, na.rm = TRUE),
                                          var  = ~var(.x, na.rm = TRUE),
                                          cv   = ~sd(.x, na.rm = TRUE) / mean(.x, na.rm = TRUE) * 100),
                                     .names = "{.col}_{.fn}"),
                        .groups = "drop")

summary_stats_long <- summary_stats %>%
    pivot_longer(cols = -c(crop, modelname),
                             names_to = c("metric", "stat"),
                             names_sep = "_",
                             values_to = "value")

print(summary_stats_long)

write.csv(summary_stats_long, "./results/summary_statistics_by_crop_model_all.csv", row.names = FALSE)
```

## Plotting distributions
```{r}
df_new <- df_no_outliers %>%
  filter(crop == "wheat", modelname == "SegNet")

ggplot(df_new, aes(x = IoU)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Model IoU", x = "IoU", y = "Frequency") +
  theme_minimal()
```

They mainly follow a left skewed distribution.
For some models and metrics, the distribution were J shaded.

## Boxplots for model comparison
```{r}
# Pivot metrics to long form and create a faceted boxplot for all metrics x crops
df_long <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value")

p <- ggplot(df_long, aes(x = modelname, y = value, fill = modelname)) +
    geom_boxplot(outlier.size = 0.8) +
    facet_grid(metric ~ crop, scales = "free_y") +
    labs(title = "Model comparison across metrics and crops", x = "Model", y = "Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
                strip.text = element_text(size = 9),
                legend.position = "bottom")

# Print plot in the notebook
p

# Save combined image
ggsave("./results/boxplots_all_metrics_crops_all.png", plot = p, width = 14, height = 20, dpi = 700)

```

A lot of outliers were observed, with the huge majority in the lower range.

## Getting the best model per crop and metric
```{r}
best_models <- summary_stats_long %>%
  filter(stat == "mean") %>%
  group_by(crop, metric) %>%
  slice_max(order_by = value, n = 1, with_ties = FALSE) %>%
  ungroup()

print(best_models)

write.csv(best_models, "./results/best_models_by_crop_metric_all.csv", row.names = FALSE)
```

U-Net for wheat and SegFormer for sorghum were the best models overall.

# Ploting bar plots for the best models

## For wheat
```{r}

df_wheat_unet <- df_no_outliers %>%
  filter(crop == "wheat", modelname == "U-NET")

# Bar plot: Mean IoU by genotype
df_genotype_means <- df_wheat_unet %>%
  group_by(genotype) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE), .groups = "drop")

max_y <- max(df_genotype_means$mean_IoU, na.rm = TRUE)
upper <- ifelse(is.finite(max_y), ceiling(max_y * 20) / 20, 1)  # round up to nearest 0.05
breaks_seq <- seq(0, upper, by = 0.05)

p_genotype <- df_genotype_means %>%
  ggplot(aes(x = factor(genotype), y = mean_IoU, fill = factor(genotype))) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper), expand = c(0, 0.02)) +
  labs(title = "Mean IoU by Genotype (Wheat, U-NET)", x = "Genotype", y = "Mean IoU") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Bar plot: Mean IoU by collectiondate
p_collectiondate <- df_wheat_unet %>%
  group_by(collectiondate) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = factor(collectiondate), y = mean_IoU, fill = factor(collectiondate))) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper), expand = c(0, 0.02)) +
  labs(title = "Mean IoU by Collection Date (Wheat, U-NET)", x = "Collection Date", y = "Mean IoU") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Combine plots side-by-side using patchwork and save
combined <- p_genotype + p_collectiondate + plot_layout(ncol = 2) +
  plot_annotation(title = "Wheat (U-NET): Mean IoU by Genotype and Collection Date")

print(combined)
ggsave("./results/mean_IoU_wheat_unet_combined_all.png", plot = combined, width = 12, height = 6, dpi = 700)

```

Genotypes 29 and 36 presented the lowest IoU values.
The collection dates were similarly performing.

## For sorghum
```{r}

df_sorghum_segformer <- df_no_outliers %>%
  filter(crop == "sorghum", modelname == "SegFormer")

# Bar plot: Mean IoU by genotype
df_genotype_means <- df_sorghum_segformer %>%
  group_by(genotype) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE), .groups = "drop")

max_y <- max(df_genotype_means$mean_IoU, na.rm = TRUE)
upper <- ifelse(is.finite(max_y), ceiling(max_y * 20) / 20, 1)  # round up to nearest 0.05
breaks_seq <- seq(0, upper, by = 0.05)

p_genotype <- df_genotype_means %>%
  ggplot(aes(x = factor(genotype), y = mean_IoU, fill = factor(genotype))) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper), expand = c(0, 0.02)) +
  labs(title = "Mean IoU by Genotype (Sorghum, SegFormer)", x = "Genotype", y = "Mean IoU") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Bar plot: Mean IoU by collectiondate
p_collectiondate <- df_sorghum_segformer %>%
  group_by(collectiondate) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(collectiondate), y = mean_IoU, fill = factor(collectiondate))) +
  geom_bar(stat = "identity") +
  labs(title = "Mean IoU by Collection Date (Sorghum, SegFormer)", x = "Collection Date", y = "Mean IoU") +
  theme_minimal() +
  theme(legend.position = "none")

# Combine plots side-by-side using patchwork and save
combined <- p_genotype + p_collectiondate + plot_layout(ncol = 2) +
  plot_annotation(title = "Sorghum (SegFormer): Mean IoU by Genotype and Collection Date")

print(combined)
ggsave("./results/mean_IoU_sorghum_segformer_combined_all.png", plot = combined, width = 12, height = 6, dpi = 700)

```

The genotypes presented a great variability in IoU values, with genotypes 11, 15, 25, 39, 44, 65, 68, and 77 showing the lowest performance.
The collection dates were similarly performing.