# Load necessary libraries
```{r}
library(tidyverse)
library(reshape2)
library(nortest)
library(car)
library(ARTool)
library(betareg)
library(WRS2)
library(tidyr)
```

# Data Import
```{r}
df <- read.csv("./results/model_evaluation_results.csv")
head(df)
```

# Descriptive Statistics

## Data summary
```{r}
summary(df)
```

## Setting variables

```{r}
crops <- c("wheat", "sorghum")
model_name <- c("DeepLabV3Plus", "U-NET", "SegNet", "SegFormer")
metrics <- c("IoU", "Dice", "Precision", "Recall", "F1")
```

## Getting Mean, Max, Min, var and cv.
```{r}
# Compute summary statistics by crop and model
summary_stats <- df %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    group_by(crop, modelname) %>%
    summarise(across(all_of(metrics),
                                     list(mean = ~mean(.x, na.rm = TRUE),
                                                max  = ~max(.x, na.rm = TRUE),
                                                min  = ~min(.x, na.rm = TRUE),
                                                var  = ~var(.x, na.rm = TRUE),
                                                cv   = ~sd(.x, na.rm = TRUE) / mean(.x, na.rm = TRUE) * 100),
                                     .names = "{.col}_{.fn}"),
                        .groups = "drop")

summary_stats_long <- summary_stats %>%
    pivot_longer(cols = -c(crop, modelname),
                             names_to = c("metric", "stat"),
                             names_sep = "_",
                             values_to = "value")

print(summary_stats_long)

write.csv(summary_stats_long, "./results/summary_statistics_by_crop_model.csv", row.names = FALSE)
```

## Plotting distributions
```{r}
df_new <- df %>%
  filter(crop == "wheat", modelname == "SegNet")

ggplot(df_new, aes(x = IoU)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Model IoU", x = "IoU", y = "Frequency") +
  theme_minimal()
```

They mainly follow a left skewed distribution.
For some models, the distribution were J shaded.
For Hausdorff and MAE, the distribution were right skewed.

## Boxplots for model comparison
```{r}
# Pivot metrics to long form and create a faceted boxplot for all metrics x crops
df_long <- df %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value")

p <- ggplot(df_long, aes(x = modelname, y = value, fill = modelname)) +
    geom_boxplot(outlier.size = 0.8) +
    facet_grid(metric ~ crop, scales = "free_y") +
    labs(title = "Model comparison across metrics and crops", x = "Model", y = "Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
                strip.text = element_text(size = 9),
                legend.position = "bottom")

# Save combined image
ggsave("./results/boxplots_all_metrics_crops.png", plot = p, width = 14, height = 20, dpi = 700)

# Print plot in the notebook
p
```

A lot of outliers were observed, with the huge majority in the lower range.

## Anderson-Darling test for normality
```{r}
normality_test_results <- df %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(metrics), 
               names_to = "metric", 
               values_to = "value") %>%
  filter(!is.na(value), !is.infinite(value), !is.nan(value)) %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    n = n(),
    n_unique = n_distinct(value),
    ad_statistic = ifelse(
      n >= 7 & n_unique >= 3,  # More stringent requirements
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$statistic)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    ad_p_value = ifelse(
      n >= 7 & n_unique >= 3,
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$p.value)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    .groups = "drop"
  )

print(normality_test_results)
write.csv(normality_test_results, "./results/anderson_darling_by_crop_model_metric.csv", row.names = FALSE)
```

Not normal at all, with p-values strongly below 0.05.

## Levene's test for homogeneity of variances
```{r}
# Levene test for homogeneity of variances (compare by genotype and/or collectiondate)
grouping_vars <- c("genotype", "collectiondate")  # change or extend as needed

levene_test_results <- df %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value") %>%
    group_by(crop, metric) %>%
    group_modify(~ {
        df_group <- .x
        purrr::map_dfr(grouping_vars, function(gvar) {
            if (!gvar %in% names(df_group)) {
                return(tibble::tibble(grouping = gvar,
                                      levene_statistic = NA_real_,
                                      levene_p_value = NA_real_))
            }
            df_sub <- df_group %>%
                filter(!is.na(.data[[gvar]]), !is.na(value), !is.infinite(value), !is.nan(value))
            
            # Convert grouping variable to factor
            df_sub[[gvar]] <- factor(df_sub[[gvar]])
            
            if (length(unique(df_sub[[gvar]])) < 2) {
                return(tibble::tibble(grouping = gvar,
                                      levene_statistic = NA_real_,
                                      levene_p_value = NA_real_))
            }
            test <- tryCatch({
                # Create formula for Levene's test
                formula_str <- paste("value ~", gvar)
                car::leveneTest(as.formula(formula_str), data = df_sub, center = median)
            }, error = function(e) {
                message("Error in Levene test for ", gvar, ": ", e$message)
                return(NULL)
            })
            
            if (is.null(test) || nrow(test) == 0) {
                tibble::tibble(grouping = gvar,
                               levene_statistic = NA_real_,
                               levene_p_value = NA_real_)
            } else {
                tibble::tibble(grouping = gvar,
                               levene_statistic = as.numeric(test[1, "F value"]),
                               levene_p_value = as.numeric(test[1, "Pr(>F)"]))
            }
        })
    }, .keep = TRUE) %>%
    ungroup() %>%
    select(crop, metric, grouping, levene_statistic, levene_p_value)

print(levene_test_results)
write.csv(levene_test_results, "./results/levene_by_crop_metric.csv", row.names = FALSE)
```

Not homogeneity in variance at all, with p-values strongly below 0.05.

## Q-Q Plots
```{r}
df %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(metrics), 
               names_to = "metric", 
               values_to = "value") %>%
  filter(!is.na(value), !is.infinite(value)) %>%
  ggplot(aes(sample = value)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ metric, scales = "free") +
  theme_minimal() +
  labs(title = "Q-Q Plots")
```

## When removing outliers
```{r}
remove_outliers <- function(x, na.rm = TRUE, ...) {
    qnt <- quantile(x, probs = c(0.25, 0.75), na.rm = na.rm, ...)
    H <- 1.5 * IQR(x, na.rm = na.rm)
    y <- x
    y[x < (qnt[1] - H)] <- NA
    y[x > (qnt[2] + H)] <- NA
    return(y)
}

df_no_outliers <- df %>%
    mutate(across(all_of(metrics), ~remove_outliers(.x)))
```

### Boxplots without outliers
```{r}
# Pivot metrics to long form and create a faceted boxplot for all metrics x crops
df_long <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value")

p <- ggplot(df_long, aes(x = modelname, y = value, fill = modelname)) +
    geom_boxplot(outlier.size = 0.8) +
    facet_grid(metric ~ crop, scales = "free_y") +
    labs(title = "Model comparison across metrics and crops", x = "Model", y = "Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
                strip.text = element_text(size = 9),
                legend.position = "bottom")

p
```

Same as before, with outliers removed.

### Normality tests without outliers
```{r}
# Histogram without outliers
df_new_no_outliers <- df_no_outliers %>%
  filter(crop == "sorghum", modelname == "SegNet")

ggplot(df_new_no_outliers, aes(x = IoU)) +
  geom_histogram(binwidth = 0.05, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Model Dice without Outliers", x = "Dice", y = "Frequency") +
  theme_minimal()

# Repeat Anderson-Darling test without outliers
normality_test_no_outliers <- df_no_outliers %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(metrics), 
               names_to = "metric", 
               values_to = "value") %>%
  filter(!is.na(value), !is.infinite(value), !is.nan(value)) %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    n = n(),
    n_unique = n_distinct(value),
    ad_statistic = ifelse(
      n >= 7 & n_unique >= 3,
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$statistic)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    ad_p_value = ifelse(
      n >= 7 & n_unique >= 3,
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$p.value)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    .groups = "drop"
  )

print(normality_test_no_outliers)

grouping_vars <- c("genotype", "collectiondate")  # change or extend as needed

levene_test_results_no_outliers <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value") %>%
    group_by(crop, metric) %>%
    group_modify(~ {
        df_group <- .x
        purrr::map_dfr(grouping_vars, function(gvar) {
            if (!gvar %in% names(df_group)) {
                return(tibble::tibble(grouping = gvar,
                                      levene_statistic = NA_real_,
                                      levene_p_value = NA_real_))
            }
            df_sub <- df_group %>%
                filter(!is.na(.data[[gvar]]), !is.na(value), !is.infinite(value), !is.nan(value))
            
            # Convert grouping variable to factor
            df_sub[[gvar]] <- factor(df_sub[[gvar]])
            
            if (length(unique(df_sub[[gvar]])) < 2) {
                return(tibble::tibble(grouping = gvar,
                                      levene_statistic = NA_real_,
                                      levene_p_value = NA_real_))
            }
            test <- tryCatch({
                # Create formula for Levene's test
                formula_str <- paste("value ~", gvar)
                car::leveneTest(as.formula(formula_str), data = df_sub, center = median)
            }, error = function(e) {
                message("Error in Levene test for ", gvar, ": ", e$message)
                return(NULL)
            })
            
            if (is.null(test) || nrow(test) == 0) {
                tibble::tibble(grouping = gvar,
                               levene_statistic = NA_real_,
                               levene_p_value = NA_real_)
            } else {
                tibble::tibble(grouping = gvar,
                               levene_statistic = as.numeric(test[1, "F value"]),
                               levene_p_value = as.numeric(test[1, "Pr(>F)"]))
            }
        })
    }, .keep = TRUE) %>%
    ungroup() %>%
    select(crop, metric, grouping, levene_statistic, levene_p_value)

print(levene_test_results_no_outliers)

```

Even when removing the outliers, the data largely did not follow a normal distribution on histograms, but kept the same trends as per the Anderson-Darling test results.

# Transforming data
```{r}
df_no_outliers <- df_no_outliers %>%
  mutate(across(all_of(metrics),
                ~{
                  mx <- max(.x, na.rm = TRUE)
                  v <- mx + 1 - .x
                  v <- ifelse(is.na(v) | is.infinite(v), NA_real_, v)
                  v <- pmax(v, .Machine$double.eps)  # avoid log(0)
                  log(v)
                },
                .names = "{.col}_transformed"))

# Histogram of transformed data
df_new <- df_no_outliers %>%
  filter(crop == "wheat", modelname == "SegNet")

ggplot(df_new, aes(x = Dice_transformed)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Model IoU", x = "IoU", y = "Frequency") +
  theme_minimal()

# Run normality tests again on transformed data
# Run tests only on transformed metrics
transformed_metrics <- paste0(metrics, "_transformed")

normality_test_transformed <- df_no_outliers %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(transformed_metrics),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = sub("_transformed$", "", metric)) %>%
  filter(!is.na(value), !is.infinite(value), !is.nan(value)) %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    n = n(),
    n_unique = n_distinct(value),
    ad_statistic = ifelse(
      n >= 7 & n_unique >= 3,
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$statistic)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    ad_p_value = ifelse(
      n >= 7 & n_unique >= 3,
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$p.value)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    .groups = "drop"
  )

print(normality_test_transformed)

grouping_vars <- c("genotype", "collectiondate")  # change or extend as needed

levene_test_transformed <- df_no_outliers %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(transformed_metrics),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = sub("_transformed$", "", metric)) %>%
  group_by(crop, metric) %>%
  group_modify(~ {
    df_group <- .x
    purrr::map_dfr(grouping_vars, function(gvar) {
      if (!gvar %in% names(df_group)) {
        return(tibble::tibble(grouping = gvar,
                              levene_statistic = NA_real_,
                              levene_p_value = NA_real_))
      }
      df_sub <- df_group %>%
        filter(!is.na(.data[[gvar]]), !is.na(value), !is.infinite(value), !is.nan(value))
      
      # Convert grouping variable to factor
      df_sub[[gvar]] <- factor(df_sub[[gvar]])
      
      if (length(unique(df_sub[[gvar]])) < 2) {
        return(tibble::tibble(grouping = gvar,
                              levene_statistic = NA_real_,
                              levene_p_value = NA_real_))
      }
      test <- tryCatch({
        formula_str <- paste("value ~", gvar)
        car::leveneTest(as.formula(formula_str), data = df_sub, center = median)
      }, error = function(e) {
        message("Error in Levene test for ", gvar, ": ", e$message)
        return(NULL)
      })
      
      if (is.null(test) || nrow(test) == 0) {
        tibble::tibble(grouping = gvar,
                       levene_statistic = NA_real_,
                       levene_p_value = NA_real_)
      } else {
        tibble::tibble(grouping = gvar,
                       levene_statistic = as.numeric(test[1, "F value"]),
                       levene_p_value = as.numeric(test[1, "Pr(>F)"]))
      }
    })
  }, .keep = TRUE) %>%
  ungroup() %>%
  select(crop, metric, grouping, levene_statistic, levene_p_value)

print(levene_test_transformed)

```
The histograms are now right skewed.
The normality and heterogeneity of variance tests are the same thing as before.

# ANOVA Analysis

```{r}
# Performing traditional ANOVA
anova_results <- df_no_outliers %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value), !is.infinite(value)) %>%
  mutate(
    genotype = factor(genotype),
    collectiondate = factor(collectiondate)
  ) %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    n = n(),
    F_genotype = tryCatch({
      m <- aov(value ~ genotype * collectiondate, data = cur_data())
      s <- summary(m)
      as.numeric(s[[1]][["F value"]][1])
    }, error = function(e) NA_real_),
    p_genotype = tryCatch({
      m <- aov(value ~ genotype * collectiondate, data = cur_data())
      s <- summary(m)
      as.numeric(s[[1]][["Pr(>F)"]][1])
    }, error = function(e) NA_real_),
    F_collectiondate = tryCatch({
      m <- aov(value ~ genotype * collectiondate, data = cur_data())
      s <- summary(m)
      as.numeric(s[[1]][["F value"]][2])
    }, error = function(e) NA_real_),
    p_collectiondate = tryCatch({
      m <- aov(value ~ genotype * collectiondate, data = cur_data())
      s <- summary(m)
      as.numeric(s[[1]][["Pr(>F)"]][2])
    }, error = function(e) NA_real_),
    F_interaction = tryCatch({
      m <- aov(value ~ genotype * collectiondate, data = cur_data())
      s <- summary(m)
      as.numeric(s[[1]][["F value"]][3])
    }, error = function(e) NA_real_),
    p_interaction = tryCatch({
      m <- aov(value ~ genotype * collectiondate, data = cur_data())
      s <- summary(m)
      as.numeric(s[[1]][["Pr(>F)"]][3])
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

print(anova_results)
write.csv(anova_results, "./results/anova_by_crop_model_metric.csv", row.names = FALSE)
```

# Analysing data using ART ANOVA

```{r}
# Aligned Rank Transform ANOVA
art_anova_results <- df %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value), !is.infinite(value)) %>%
  group_by(crop, modelname, metric) %>%
  filter(n_distinct(value) > 1, n() >= 3) %>%  # Only keep groups with variance and enough data
  mutate(
    genotype = factor(genotype),
    collectiondate = factor(collectiondate)
  ) %>%
  summarise(
    n = n(),
    F_genotype = tryCatch({
      m <- art(value ~ genotype * collectiondate, data = cur_data())
      anova_table <- anova(m)
      anova_table$`F`[1]
    }, error = function(e) NA_real_),
    p_genotype = tryCatch({
      m <- art(value ~ genotype * collectiondate, data = cur_data())
      anova_table <- anova(m)
      anova_table$`Pr(>F)`[1]
    }, error = function(e) NA_real_),
    F_date = tryCatch({
      m <- art(value ~ genotype * collectiondate, data = cur_data())
      anova_table <- anova(m)
      anova_table$`F`[2]
    }, error = function(e) NA_real_),
    p_date = tryCatch({
      m <- art(value ~ genotype * collectiondate, data = cur_data())
      anova_table <- anova(m)
      anova_table$`Pr(>F)`[2]
    }, error = function(e) NA_real_),
    F_interaction = tryCatch({
      m <- art(value ~ genotype * collectiondate, data = cur_data())
      anova_table <- anova(m)
      anova_table$`F`[3]
    }, error = function(e) NA_real_),
    p_interaction = tryCatch({
      m <- art(value ~ genotype * collectiondate, data = cur_data())
      anova_table <- anova(m)
      anova_table$`Pr(>F)`[3]
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

print(art_anova_results)
write.csv(art_anova_results, "./results/art_anova_by_crop_model_metric.csv", row.names = FALSE)
```

# Beta Transformation
```{r}

df_new <- df %>%
  filter(crop == "wheat", modelname == "SegFormer")

df_new$genotype <- as.factor(df_new$genotype)
df_new$collectiondate <- as.factor(df_new$collectiondate)

#Removing 0 and 1
df_new <- df_new %>%
  filter(IoU > 0 & IoU < 1) %>%
  filter(Precision > 0 & Precision < 1)

m.betareg.1 <- betareg(IoU ~ collectiondate * genotype, data = df_new)
m.betareg.2 <- betareg(IoU ~ collectiondate + genotype, data = df_new)
summary(m.betareg.1)
summary(m.betareg.2)

library(lmtest)
lrtest(m.betareg.1, m.betareg.2)

invlogit <- function(x) {1/(1+exp(-x))}
invlogit(coef(m.betareg))

library(multcomp)
library(multcompView)
library(emmeans)
m.betareg_iou.em <- emmeans(m.betareg.1, ~ genotype | collectiondate)
cld(m.betareg_iou.em, Letters = letters, sort = FALSE)

```

# Robust ANOVA
```{r}

sorghum_genotypes <- c("8", "14", "25", "37", "55", "64", "70", "79")

df_new <- df %>%
  filter(crop == "sorghum", modelname == "SegFormer", !(genotype %in% sorghum_genotypes))

df_new <- df %>%
  filter(crop == "wheat", modelname == "SegFormer")

df_new$genotype <- as.factor(df_new$genotype)
df_new$collectiondate <- as.factor(df_new$collectiondate)

df_new <- df_new %>%
  filter(IoU > 0 & IoU < 1)

table(df_new$genotype, df_new$collectiondate)

anova <- aov(IoU ~ genotype * collectiondate, data = df_new)
Anova(anova, type = "III")

t <- WRS2::t2way(IoU ~ genotype * collectiondate, data = df_new, tr = 0.2)
print(t)

```