# Load necessary libraries
```{r}
library(tidyverse)
library(patchwork)
```

# Data Import
```{r}
df <- read.csv("./results/model_evaluation_results.csv")
head(df)
```

# Descriptive Statistics

## Setting variables

```{r}
crops <- c("wheat", "sorghum", "corn")
model_name <- c("DeepLabV3Plus", "U-NET", "SegNet", "SegFormer", "MaskFormer")
metrics <- c("IoU", "Precision", "Recall", "F1")
```

## Data summary
```{r}
summary(df)
```

## When removing outliers
```{r}
remove_outliers <- function(x, na.rm = TRUE, ...) {
    qnt <- quantile(x, probs = c(0.25, 0.75), na.rm = na.rm, ...)
    H <- 1.5 * IQR(x, na.rm = na.rm)
    y <- x
    y[x < (qnt[1] - H)] <- NA
    y[x > (qnt[2] + H)] <- NA
    return(y)
}

df_no_outliers <- df %>%
    mutate(across(all_of(metrics), ~remove_outliers(.x))) %>%
    select(crop, modelname, all_of(metrics), genotype, collectiondate) %>%
    drop_na()

summary(df_no_outliers)

table(df_no_outliers$crop, df_no_outliers$modelname)
```

## Getting Mean, Max, Min, var and cv.
```{r}
# Compute summary statistics by crop and model
summary_stats <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    group_by(crop, modelname) %>%
    summarise(across(all_of(metrics),
                                     list(mean = ~mean(.x, na.rm = TRUE),
                                          max  = ~max(.x, na.rm = TRUE),
                                          min  = ~min(.x, na.rm = TRUE),
                                          var  = ~var(.x, na.rm = TRUE),
                                          cv   = ~sd(.x, na.rm = TRUE) / mean(.x, na.rm = TRUE) * 100),
                                     .names = "{.col}_{.fn}"),
                        .groups = "drop")

summary_stats_long <- summary_stats %>%
    pivot_longer(cols = -c(crop, modelname),
                             names_to = c("metric", "stat"),
                             names_sep = "_",
                             values_to = "value")

print(summary_stats_long)

write.csv(summary_stats_long, "./results/summary_statistics_by_crop_model_all.csv", row.names = FALSE)
```

## Summary Stats Short
```{r}
# Short summary table with mean and CV (%) for selected metrics
summary_short <- df_no_outliers %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  group_by(crop, modelname) %>%
  summarise(
    IoUNew = mean(IoU, na.rm = TRUE),
    IoUCV = (sd(IoU, na.rm = TRUE) / mean(IoU, na.rm = TRUE)) * 100,
    PrecisionNew = mean(Precision, na.rm = TRUE),
    PrecisionCV = (sd(Precision, na.rm = TRUE) / mean(Precision, na.rm = TRUE)) * 100,
    RecallNew = mean(Recall, na.rm = TRUE),
    RecallCV = (sd(Recall, na.rm = TRUE) / mean(Recall, na.rm = TRUE)) * 100,
    F1New = mean(F1, na.rm = TRUE),
    F1CV = (sd(F1, na.rm = TRUE) / mean(F1, na.rm = TRUE)) * 100,
    .groups = "drop"
  )
  select(
    Crop      = crop,
    Model     = modelname,
    IoUNew,
    IoUCV,
    PrecisionNew,
    PrecisionCV,
    RecallNew,
    RecallCV,
    F1New,
    F1CV
  ) %>%
  # rounding for presentation
  mutate(
    IoUNew = round(IoUNew, 3),
    PrecisionNew = round(PrecisionNew, 3),
    RecallNew = round(RecallNew, 3),
    F1New = round(F1New, 3),
    IoUCV = round(IoUCV, 2),
    PrecisionCV = round(PrecisionCV, 2),
    RecallCV = round(RecallCV, 2),
    F1CV = round(F1CV, 2)
  ) %>%
  arrange(Crop, Model)

# Print and save
print(summary_short)
write.csv(summary_short, "./results/summary_stats_short_by_crop_model.csv", row.names = FALSE)
```

## Plotting distributions
```{r}
df_new <- df_no_outliers %>%
  filter(crop == "wheat", modelname == "SegNet")

ggplot(df_new, aes(x = IoU)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Model IoU", x = "IoU", y = "Frequency") +
  theme_minimal()
```

They mainly follow a left skewed distribution.
For some models and metrics, the distribution were J shaded.

## Boxplots for model comparison
```{r}
# Pivot metrics to long form and create a faceted boxplot for all metrics x crops
df_long <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value")

p <- ggplot(df_long, aes(x = modelname, y = value, fill = modelname)) +
    geom_boxplot(outlier.size = 0.8) +
    facet_grid(metric ~ crop, scales = "free_y") +
    labs(title = "Model comparison across metrics and crops", x = "Model", y = "Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
                strip.text = element_text(size = 9),
                legend.position = "bottom")

# Print plot in the notebook
p

# Save combined image
ggsave("./results/boxplots_all_metrics_crops_all.png", plot = p, width = 14, height = 20, dpi = 700)

```

A lot of outliers were observed, with the huge majority in the lower range.

## Getting the best model per crop and metric
```{r}
best_models <- summary_stats_long %>%
  filter(stat == "mean") %>%
  group_by(crop, metric) %>%
  slice_max(order_by = value, n = 1, with_ties = FALSE) %>%
  ungroup()

print(best_models)

write.csv(best_models, "./results/best_models_by_crop_metric_all.csv", row.names = FALSE)
```

U-Net for wheat and SegFormer for sorghum were the best models overall.

# Ploting bar plots for the best models

```{r}
purdue_colors <- c(
  "black" = "#000000",
  "gold" = "#CEB888",
  "white" = "#FFFFFF",
  "steel" = "#555960"
)
# For more categories, use gold and black alternately, or use a custom palette:
purdue_palette <- c("#CEB888", "#555960", "#B6A76A", "#7A6A53", "#F3E5AB", "#A8996E", "#6E6E6E", "#E1C16E")
```

## For wheat
```{r}

df_wheat_unet <- df_no_outliers %>%
  filter(crop == "wheat", modelname == "U-NET")

# Bar plot: Mean IoU by genotype (use SD for error bars)
df_genotype_stats <- df_wheat_unet %>%
  group_by(genotype) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE),
            sd = sd(IoU, na.rm = TRUE),
            n = sum(!is.na(IoU)),
            lower = pmax(0, mean_IoU - sd),
            upper = mean_IoU + sd,
            .groups = "drop")

max_y <- max(df_genotype_stats$upper, na.rm = TRUE)
upper_lim <- ifelse(is.finite(max_y), ceiling(max_y * 20) / 20, 1)  # round up to nearest 0.05
breaks_seq <- seq(0, upper_lim, by = 0.05)

p_genotype <- df_genotype_stats %>%
  ggplot(aes(x = factor(genotype), y = mean_IoU, fill = factor(genotype))) +
  geom_col(color = purdue_colors["black"], width = 0.7, show.legend = FALSE) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = purdue_colors["black"], linewidth = 1) +
  scale_fill_manual(values = rep(purdue_palette, length.out = nrow(df_genotype_stats))) +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper_lim), expand = c(0, 0.02)) +
  labs(title = "Genotype", x = "Genotype", y = "Mean IoU") +
  theme_minimal(base_size = 16) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = purdue_colors["black"], size = 12),
    axis.text.y = element_text(color = purdue_colors["black"], size = 12),
    axis.title = element_text(color = purdue_colors["black"], size = 14, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 18, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.3),
    panel.grid.minor = element_blank()
  )

print(p_genotype)

df_date_stats <- df_wheat_unet %>%
  group_by(collectiondate) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE),
            sd = sd(IoU, na.rm = TRUE),
            n = sum(!is.na(IoU)),
            lower = pmax(0, mean_IoU - sd),
            upper = mean_IoU + sd,
            .groups = "drop")

# Use same y-scale as genotype plot
max_y_date <- max(df_date_stats$upper, na.rm = TRUE)
upper_lim <- ifelse(is.finite(max(df_genotype_stats$upper, max_y_date)), ceiling(max(df_genotype_stats$upper, max_y_date) * 20) / 20, 1)
breaks_seq <- seq(0, upper_lim, by = 0.05)

p_collectiondate <- df_date_stats %>%
  ggplot(aes(x = factor(collectiondate), y = mean_IoU, fill = factor(collectiondate))) +
  geom_col(color = purdue_colors["black"], width = 0.7, show.legend = FALSE) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = purdue_colors["black"], linewidth = 1) +
  scale_fill_manual(values = rep(purdue_palette, length.out = nrow(df_date_stats))) +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper_lim), expand = c(0, 0.02)) +
  labs(title = "Collection Date", x = "Collection Date", y = "Mean IoU") +
  theme_minimal(base_size = 16) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = purdue_colors["black"], size = 12),
    axis.text.y = element_text(color = purdue_colors["black"], size = 12),
    axis.title = element_text(color = purdue_colors["black"], size = 14, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 18, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.3),
    panel.grid.minor = element_blank()
  )

print(p_collectiondate)

# Combine plots side-by-side using patchwork and save
combined <- p_genotype + p_collectiondate + plot_layout(ncol = 2) +
  plot_annotation(
    title = "Wheat (U-NET): Mean IoU by Genotype and Collection Date",
    theme = theme(
      plot.title = element_text(color = purdue_colors["black"], size = 20, face = "bold", hjust = 0.5),
      plot.background = element_rect(fill = purdue_colors["white"], color = NA)
    )
  )

print(combined)
ggsave("./results/mean_IoU_wheat_unet_combined_all.png", plot = combined, width = 20, height = 6, dpi = 700)

```

### For sorghum
```{r}

df_sorghum_segformer <- df_no_outliers %>%
  filter(crop == "sorghum", modelname == "SegFormer")

# Bar plot: Mean IoU by genotype (use SD for error bars)
df_genotype_stats <- df_sorghum_segformer %>%
  group_by(genotype) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE),
            sd = sd(IoU, na.rm = TRUE),
            n = sum(!is.na(IoU)),
            lower = pmax(0, mean_IoU - sd),
            upper = mean_IoU + sd,
            .groups = "drop")

max_y <- max(df_genotype_stats$upper, na.rm = TRUE)
upper_lim <- ifelse(is.finite(max_y), ceiling(max_y * 20) / 20, 1)  # round up to nearest 0.05
breaks_seq <- seq(0, upper_lim, by = 0.05)

p_genotype <- df_genotype_stats %>%
  ggplot(aes(x = factor(genotype), y = mean_IoU, fill = factor(genotype))) +
  geom_col(color = purdue_colors["black"], width = 0.7, show.legend = FALSE) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = purdue_colors["black"], linewidth = 1) +
  scale_fill_manual(values = rep(purdue_palette, length.out = nrow(df_genotype_stats))) +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper_lim), expand = c(0, 0.02)) +
  labs(title = "Genotype", x = "Genotype", y = "Mean IoU") +
  theme_minimal(base_size = 16) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = purdue_colors["black"], size = 12),
    axis.text.y = element_text(color = purdue_colors["black"], size = 12),
    axis.title = element_text(color = purdue_colors["black"], size = 14, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 18, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.3),
    panel.grid.minor = element_blank()
  )

print(p_genotype)

# Bar plot: Mean IoU by collection date (use SD for error bars)
df_date_stats <- df_sorghum_segformer %>%
  group_by(collectiondate) %>%
  summarise(mean_IoU = mean(IoU, na.rm = TRUE),
            sd = sd(IoU, na.rm = TRUE),
            n = sum(!is.na(IoU)),
            lower = pmax(0, mean_IoU - sd),
            upper = mean_IoU + sd,
            .groups = "drop")

max_y_date <- max(df_date_stats$upper, na.rm = TRUE)
upper_lim <- ifelse(is.finite(max(df_genotype_stats$upper, max_y_date)), ceiling(max(df_genotype_stats$upper, max_y_date) * 20) / 20, 1)
breaks_seq <- seq(0, upper_lim, by = 0.05)

p_collectiondate <- df_date_stats %>%
  ggplot(aes(x = factor(collectiondate), y = mean_IoU, fill = factor(collectiondate))) +
  geom_col(color = purdue_colors["black"], width = 0.7, show.legend = FALSE) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = purdue_colors["black"], linewidth = 1) +
  scale_fill_manual(values = rep(purdue_palette, length.out = nrow(df_date_stats))) +
  scale_y_continuous(breaks = breaks_seq, limits = c(0, upper_lim), expand = c(0, 0.02)) +
  labs(title = "Collection Date", x = "Collection Date", y = "Mean IoU") +
  theme_minimal(base_size = 16) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, color = purdue_colors["black"], size = 12),
    axis.text.y = element_text(color = purdue_colors["black"], size = 12),
    axis.title = element_text(color = purdue_colors["black"], size = 14, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 18, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.3),
    panel.grid.minor = element_blank()
  )

print(p_collectiondate)

# Combine plots side-by-side using patchwork and save
combined <- p_genotype + p_collectiondate + plot_layout(ncol = 2) +
  plot_annotation(title = "Sorghum (SegFormer): Mean IoU by Genotype and Collection Date",
                  theme = theme(
                      plot.title = element_text(color = purdue_colors["black"], size = 20, face = "bold", hjust = 0.5),
                      plot.background = element_rect(fill = purdue_colors["white"], color = NA)
                    )
                  )

print(combined)
ggsave("./results/mean_IoU_sorghum_segformer_combined_all.png", plot = combined, width = 20, height = 6, dpi = 700)

```

The genotypes presented a great variability in IoU values, with genotypes 11, 15, 25, 39, 44, 65, 68, and 77 showing the lowest performance.
The collection dates were similarly performing.

### Comparing Same individuals
```{r}

df_individuals <- read.csv("./results/model_evaluation_results_individuals.csv")
head(df_individuals)

df_no_outliers_individuals <- df_individuals %>%
    mutate(across(all_of(metrics), ~remove_outliers(.x))) %>%
    filter(`augmented.` == "no") %>%
    select(crop, modelname, all_of(metrics), genotype, collectiondate, individual_id) %>%
    drop_na()

head(df_no_outliers_individuals)
summary(df_no_outliers_individuals)

# Summarize IoU for each individual across collection dates
individual_variance <- df_no_outliers_individuals %>%
  group_by(crop, modelname, genotype, individual_id) %>%
  summarise(
    n_dates = n_distinct(collectiondate),
    mean_IoU = mean(IoU, na.rm = TRUE),
    sd_IoU = sd(IoU, na.rm = TRUE),
    min_IoU = min(IoU, na.rm = TRUE),
    max_IoU = max(IoU, na.rm = TRUE),
    .groups = "drop"
  )

print(individual_variance)

# Prepare common x/y axis breaks so both histograms share the same ticks
binwidth <- 0.01

sd_w <- individual_variance %>% filter(crop == "wheat", modelname == "U-NET") %>% pull(sd_IoU)
sd_s <- individual_variance %>% filter(crop == "sorghum", modelname == "SegFormer") %>% pull(sd_IoU)

upper_x_raw <- max(c(sd_w, sd_s), na.rm = TRUE)
upper_x <- ifelse(is.finite(upper_x_raw) && upper_x_raw > 0,
                  ceiling(upper_x_raw / binwidth) * binwidth,
                  binwidth)
breaks_x <- seq(0, upper_x, by = binwidth)

# helper to get bin counts robustly
bin_counts <- function(vals, breaks) {
  if (length(vals) == 0) return(rep(0, length(breaks) - 1))
  as.numeric(table(cut(vals, breaks = breaks, include.lowest = TRUE)))
}
counts_w <- bin_counts(sd_w, breaks_x)
counts_s <- bin_counts(sd_s, breaks_x)
max_count <- max(c(counts_w, counts_s), na.rm = TRUE)
y_step <- ifelse(max_count <= 10, 1, ceiling(max_count / 5))
max_y <- ifelse(is.finite(max_count) && max_count > 0, ceiling(max_count / y_step) * y_step, 1)
breaks_y <- seq(0, max_y, by = y_step)

# Wheat (U-NET) SD histogram with common ticks
p_wheat_sd <- individual_variance %>%
  filter(crop == "wheat", modelname == "U-NET") %>%
  ggplot(aes(x = sd_IoU)) +
  geom_histogram(binwidth = binwidth, fill = purdue_colors["steel"], color = purdue_colors["black"], alpha = 0.8) +
  scale_x_continuous(breaks = breaks_x, limits = c(0, upper_x), expand = c(0, 0.01)) +
  scale_y_continuous(breaks = breaks_y, limits = c(0, max_y), expand = c(0, 0.02)) +
  labs(title = "(Wheat, U-NET)", x = "SD of IoU", y = "Count") +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text = element_text(color = purdue_colors["black"], size = 11),
    axis.title = element_text(color = purdue_colors["black"], size = 12, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.25),
    panel.grid.minor = element_blank()
  )

print(p_wheat_sd)

# Sorghum (SegFormer) SD histogram with common ticks
p_sorghum_sd <- individual_variance %>%
  filter(crop == "sorghum", modelname == "SegFormer") %>%
  ggplot(aes(x = sd_IoU)) +
  geom_histogram(binwidth = binwidth, fill = purdue_colors["steel"], color = purdue_colors["black"], alpha = 0.8) +
  scale_x_continuous(breaks = breaks_x, limits = c(0, upper_x), expand = c(0, 0.01)) +
  scale_y_continuous(breaks = breaks_y, limits = c(0, max_y), expand = c(0, 0.02)) +
  labs(title = "(Sorghum, SegFormer)", x = "SD of IoU", y = "Count") +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text = element_text(color = purdue_colors["black"], size = 11),
    axis.title = element_text(color = purdue_colors["black"], size = 12, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.25),
    panel.grid.minor = element_blank()
  )

print(p_sorghum_sd)

# Combine side-by-side and save
combined_sd <- p_wheat_sd + p_sorghum_sd + plot_layout(ncol = 2) +
  plot_annotation(title = "Distribution of IoU Variance (SD) Across Individuals: Wheat (U-NET) vs Sorghum (SegFormer)",
                  theme = theme(
                      plot.title = element_text(color = purdue_colors["black"], size = 18, face = "bold", hjust = 0.5),
                      plot.background = element_rect(fill = purdue_colors["white"], color = NA)
                    )
                  )

print(combined_sd)

ggsave("./results/IoU_SD_histograms_wheat_unet_sorghum_segformer_combined.png",
       plot = combined_sd, width = 14, height = 6, dpi = 600)

```

## Compare genotypes within each crop/model and get min/max mean per metric

```{r}
# Select the models
target_models <- tribble(
  ~crop,     ~modelname,
  "corn",    "SegFormer",
  "wheat",   "U-NET",
  "sorghum", "SegFormer"
)

# Make sure crop and modelname cases match your df_no_outliers
df_sel <- df_no_outliers %>%
  mutate(crop = trimws(tolower(as.character(crop))),
         modelname = trimws(as.character(modelname))) %>%
  # keep only selected pairs
  inner_join(target_models %>% mutate(crop = trimws(tolower(crop))),
             by = c("crop", "modelname"))

# Compute per-genotype mean, sd and CV (%) for each metric
df_genotype_mean <- df_sel %>%
  group_by(crop, modelname, genotype) %>%
  summarise(
    IoU_mean = mean(IoU, na.rm = TRUE),
    Precision_mean = mean(Precision, na.rm = TRUE),
    Recall_mean = mean(Recall, na.rm = TRUE),
    F1_mean = mean(F1, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Pivot CV columns and compute extremes per crop/model/metric
metrics_mean <- c("IoU_mean", "Precision_mean", "Recall_mean", "F1_mean")

df_extremes <- df_genotype_mean %>%
  pivot_longer(cols = all_of(metrics_mean), names_to = "metric", values_to = "Mean") %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    min_mean = if (all(is.na(Mean))) NA_real_ else min(Mean, na.rm = TRUE),
    min_genotype = if (all(is.na(Mean))) NA else genotype[which.min(replace_na(Mean, Inf))],

    max_mean = if (all(is.na(Mean))) NA_real_ else max(Mean, na.rm = TRUE),
    max_genotype = if (all(is.na(Mean))) NA else genotype[which.max(replace_na(Mean, -Inf))],

    .groups = "drop"
  ) %>%
  # clean metric names for readability
  mutate(metric = case_when(
    metric == "IoU_mean" ~ "IoU Mean",
    metric == "Precision_mean" ~ "Precision Mean",
    metric == "Recall_mean" ~ "Recall Mean",
    metric == "F1_mean" ~ "F1 Mean",
    TRUE ~ metric
  )) %>%
  arrange(crop, modelname, metric)

# Round numerical CV columns for presentation
df_extremes <- df_extremes %>%
  mutate(
    min_mean = round(min_mean, 2),
    max_mean = round(max_mean, 2)
  )

# Print and save
print(df_extremes)
write.csv(df_extremes, "./results/genotype_mean_extremes_by_crop_model.csv", row.names = FALSE)
```

## Mean IoU by crop per collection date (grouped bar chart with error bars)

```{r}
# Select the models
target_models <- tribble(
  ~crop,     ~modelname,
  "corn",    "SegFormer",
  "wheat",   "U-NET",
  "sorghum", "SegFormer"
)

# Ensure collectiondate is a character and ordered 1..6
df_sel_dates <- df_no_outliers %>%
  mutate(crop = trimws(tolower(as.character(crop))),
         modelname = trimws(as.character(modelname))) %>%
  # keep only selected pairs
  inner_join(target_models %>% mutate(crop = trimws(tolower(crop))),
             by = c("crop", "modelname")) %>%
  mutate(collectiondate = as.character(collectiondate),
         collectiondate = trimws(collectiondate))

df_summary_date <- df_sel_dates %>%
  group_by(crop, collectiondate) %>%
  summarise(
    mean_IoU = mean(IoU, na.rm = TRUE),
    sd_IoU   = sd(IoU, na.rm = TRUE),
    n        = sum(!is.na(IoU)),
    lower    = pmax(0, mean_IoU - sd_IoU),
    upper    = pmin(1, mean_IoU + sd_IoU),
    .groups = "drop"
  )

# Make sure crop is a factor with correct levels for color mapping
df_summary_date$crop <- factor(df_summary_date$crop, levels = c("wheat", "sorghum", "corn"))

# Order collection dates including 6
df_summary_date$collectiondate <- factor(df_summary_date$collectiondate, levels = c("1","2","3","4","5","6"))

crops <- unique(as.character(df_summary_date$crop))

purdue_colors <- c(
  "black" = "#000000",
  "gold" = "#CEB888",
  "steel" = "#555960"
)

color_vals <- setNames(purdue_colors[seq_len(length(crops))], crops)

label_vals <- case_when(
  crops == "wheat" ~ "Wheat (U-NET)",
  crops == "sorghum" ~ "Sorghum (SegFormer)",
  crops == "corn" ~ "Corn (SegFormer)",
  TRUE                 ~ crops
)
names(label_vals) <- crops
# Dodge width
dodge_w <- 0.8

p_date_crop <- ggplot(df_summary_date, aes(x = collectiondate, y = mean_IoU, color = crop, group = crop)) +
  geom_line(size = 2, alpha = 0.9) +
  geom_point(size = 3.5, shape = 21, fill = "black", stroke = 1.5) +
  scale_color_manual(
    values = color_vals,
    labels = label_vals,
    breaks = names(label_vals),
    name = "Crop") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 20), limits = c(0, 1), expand = c(0, 0.01)) +
  labs(x = "Collection Date",
       y = "Mean IoU ± SD") +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text.x = element_text(color = purdue_colors["black"], size = 12),
    axis.text.y = element_text(color = purdue_colors["black"], size = 12),
    axis.title = element_text(color = purdue_colors["black"], size = 13, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 16, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(color = purdue_colors["black"], size = 12, face = "bold"),
    legend.text = element_text(color = purdue_colors["black"], size = 11),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.3),
    panel.grid.minor = element_blank()
  )

print(p_date_crop)
ggsave("./results/mean_IoU_by_crop_by_date.png", plot = p_date_crop, width = 8, height = 10, dpi = 600)
```