# Load necessary libraries
```{r}
library(tidyverse)
library(reshape2)
library(nortest)
```

# Data Import
```{r}
df <- read.csv("./results/model_evaluation_results.csv")
head(df)
```

# Descriptive Statistics

## Data summary
```{r}
summary(df)
```

## Setting variables

```{r}
crops <- c("wheat", "sorghum")
model_name <- c("DeepLabV3Plus", "U-NET", "SegNet", "SegFormer")
metrics <- c("IoU", "Dice", "Precision", "Recall", "F1", "MAE", "Hausdorff", "PixelAccuracy")
```

## Getting Mean, Max, Min, var and cv.
```{r}
# Compute summary statistics by crop and model
summary_stats <- df %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    group_by(crop, modelname) %>%
    summarise(across(all_of(metrics),
                                     list(mean = ~mean(.x, na.rm = TRUE),
                                                max  = ~max(.x, na.rm = TRUE),
                                                min  = ~min(.x, na.rm = TRUE),
                                                var  = ~var(.x, na.rm = TRUE),
                                                cv   = ~sd(.x, na.rm = TRUE) / mean(.x, na.rm = TRUE) * 100),
                                     .names = "{.col}_{.fn}"),
                        .groups = "drop")

# Optional: a long/tidy version (metric, stat, value)
summary_stats_long <- summary_stats %>%
    pivot_longer(cols = -c(crop, modelname),
                             names_to = c("metric", "stat"),
                             names_sep = "_",
                             values_to = "value")

print(summary_stats_long)

write.csv(summary_stats_long, "./results/summary_statistics_by_crop_model.csv", row.names = FALSE)
```

## Plotting distributions
```{r}

df_new <- df %>%
  filter(crop == "wheat", modelname == "SegNet")

ggplot(df_new, aes(x = Hausdorff)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Model Dice", x = "Dice", y = "Frequency") +
  theme_minimal()
```

They mainly follow a left skewed distribution.
For some models, the distribution were J shaded.
For Hausdorff and MAE, the distribution were right skewed.

## Boxplots for model comparison
```{r}
# Pivot metrics to long form and create a faceted boxplot for all metrics x crops
df_long <- df %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value")

p <- ggplot(df_long, aes(x = modelname, y = value, fill = modelname)) +
    geom_boxplot(outlier.size = 0.8) +
    facet_grid(metric ~ crop, scales = "free_y") +
    labs(title = "Model comparison across metrics and crops", x = "Model", y = "Value") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
                strip.text = element_text(size = 9),
                legend.position = "bottom")

# Save combined image
ggsave("./results/boxplots_all_metrics_crops.png", plot = p, width = 14, height = 20, dpi = 700)

# Print plot in the notebook
p
```

## Pearson Correlation Analysis
```{r}
# Compute correlation matrix for selected metrics
correlation_data <- df %>%
  select(all_of(metrics))
cor_matrix <- cor(correlation_data, use = "complete.obs", method = "pearson")
print(cor_matrix)

# Visualize correlation matrix
cor_melted <- melt(cor_matrix)
ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", value)), size = 3) +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                                             limit = c(-1, 1), name="Pearson\nCorrelation") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                                                     size = 10, hjust = 1)) +
    coord_fixed()

ggsave("./results/correlation_matrix.png", plot = last_plot(), width = 10, height = 8, dpi = 500)

```

## Anderson-Darling test for normality
```{r}
normality_test_results <- df %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  pivot_longer(cols = all_of(metrics), 
               names_to = "metric", 
               values_to = "value") %>%
  filter(!is.na(value), !is.infinite(value), !is.nan(value)) %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    n = n(),
    n_unique = n_distinct(value),
    ad_statistic = ifelse(
      n >= 7 & n_unique >= 3,  # More stringent requirements
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$statistic)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    ad_p_value = ifelse(
      n >= 7 & n_unique >= 3,
      tryCatch({
        test_result <- nortest::ad.test(value)
        as.numeric(test_result$p.value)
      }, error = function(e) NA_real_),
      NA_real_
    ),
    .groups = "drop"
  )

print(normality_test_results)
write.csv(normality_test_results, "./results/anderson_darling_by_crop_model_metric.csv", row.names = FALSE)
```

## Bartlett test for homogeneity of variances
```{r}
# Bartlett test for homogeneity of variances (compare by genotype and/or collectiondate)
grouping_vars <- c("genotype", "collectiondate")  # change or extend as needed

bartlett_test_results <- df %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    pivot_longer(cols = all_of(metrics), names_to = "metric", values_to = "value") %>%
    group_by(crop, metric) %>%
    group_modify(~ {
        df_group <- .x
        purrr::map_dfr(grouping_vars, function(gvar) {
            if (!gvar %in% names(df_group)) {
                return(tibble::tibble(grouping = gvar,
                                                            bartlett_statistic = NA_real_,
                                                            bartlett_p_value = NA_real_))
            }
            df_sub <- df_group %>% filter(!is.na(.data[[gvar]]), !is.na(value))
            if (length(unique(df_sub[[gvar]])) < 2) {
                return(tibble::tibble(grouping = gvar,
                                                            bartlett_statistic = NA_real_,
                                                            bartlett_p_value = NA_real_))
            }
            test <- tryCatch(bartlett.test(df_sub$value ~ factor(df_sub[[gvar]])),
                                             error = function(e) NULL)
            if (is.null(test)) {
                tibble::tibble(grouping = gvar,
                                             bartlett_statistic = NA_real_,
                                             bartlett_p_value = NA_real_)
            } else {
                tibble::tibble(grouping = gvar,
                                             bartlett_statistic = as.numeric(test$statistic),
                                             bartlett_p_value = as.numeric(test$p.value))
            }
        })
    }, .keep = TRUE) %>%
    ungroup() %>%
    select(crop, metric, grouping, bartlett_statistic, bartlett_p_value)

print(bartlett_test_results)
write.csv(bartlett_test_results, "./results/bartlett_by_crop_metric.csv", row.names = FALSE)
```