# Load necessary libraries
```{r}
library(tidyverse)
library(patchwork)
library(car)
```

# Data Import
```{r}
df <- read.csv("./results/model_evaluation_results_rev.csv")
head(df)
```

# Descriptive Statistics

## Setting variables

```{r}
crops <- c("wheat", "sorghum", "corn")
model_name <- c("DeepLabV3Plus", "U-NET", "SegNet", "SegFormer", "MaskFormer")
metrics <- c("IoU", "Precision", "Recall", "F1")
```

## Data summary
```{r}
summary(df)
```

## When removing outliers
```{r}
remove_outliers <- function(x, na.rm = TRUE, ...) {
    qnt <- quantile(x, probs = c(0.25, 0.75), na.rm = na.rm, ...)
    H <- 1.5 * IQR(x, na.rm = na.rm)
    y <- x
    y[x < (qnt[1] - H)] <- NA
    y[x > (qnt[2] + H)] <- NA
    return(y)
}

df_no_outliers <- df %>%
    mutate(across(all_of(metrics), ~remove_outliers(.x))) %>%
    select(crop, modelname, all_of(metrics), genotype, collectiondate) %>%
    drop_na()

summary(df_no_outliers)

table(df_no_outliers$crop, df_no_outliers$modelname)
```

## Getting Mean, Max, Min, var and cv.
```{r}
# Compute summary statistics by crop and model
summary_stats <- df_no_outliers %>%
    filter(modelname %in% model_name, crop %in% crops) %>%
    group_by(crop, modelname) %>%
    summarise(across(all_of(metrics),
                                     list(mean = ~mean(.x, na.rm = TRUE),
                                          max  = ~max(.x, na.rm = TRUE),
                                          min  = ~min(.x, na.rm = TRUE),
                                          var  = ~var(.x, na.rm = TRUE),
                                          cv   = ~sd(.x, na.rm = TRUE) / mean(.x, na.rm = TRUE) * 100),
                                     .names = "{.col}_{.fn}"),
                        .groups = "drop")

summary_stats_long <- summary_stats %>%
    pivot_longer(cols = -c(crop, modelname),
                             names_to = c("metric", "stat"),
                             names_sep = "_",
                             values_to = "value")

print(summary_stats_long)

write.csv(summary_stats_long, "./results/summary_statistics_by_crop_model_all.csv", row.names = FALSE)
```

## Summary Stats Short
```{r}
# Short summary table with mean and CV (%) for selected metrics
summary_short <- df_no_outliers %>%
  filter(modelname %in% model_name, crop %in% crops) %>%
  group_by(crop, modelname) %>%
  summarise(
    IoUNew = mean(IoU, na.rm = TRUE),
    IoUCV = (sd(IoU, na.rm = TRUE) / mean(IoU, na.rm = TRUE)) * 100,
    PrecisionNew = mean(Precision, na.rm = TRUE),
    PrecisionCV = (sd(Precision, na.rm = TRUE) / mean(Precision, na.rm = TRUE)) * 100,
    RecallNew = mean(Recall, na.rm = TRUE),
    RecallCV = (sd(Recall, na.rm = TRUE) / mean(Recall, na.rm = TRUE)) * 100,
    F1New = mean(F1, na.rm = TRUE),
    F1CV = (sd(F1, na.rm = TRUE) / mean(F1, na.rm = TRUE)) * 100,
    .groups = "drop"
  )
  select(
    Crop      = crop,
    Model     = modelname,
    IoUNew,
    IoUCV,
    PrecisionNew,
    PrecisionCV,
    RecallNew,
    RecallCV,
    F1New,
    F1CV
  ) %>%
  # rounding for presentation
  mutate(
    IoUNew = round(IoUNew, 3),
    PrecisionNew = round(PrecisionNew, 3),
    RecallNew = round(RecallNew, 3),
    F1New = round(F1New, 3),
    IoUCV = round(IoUCV, 2),
    PrecisionCV = round(PrecisionCV, 2),
    RecallCV = round(RecallCV, 2),
    F1CV = round(F1CV, 2)
  ) %>%
  arrange(Crop, Model)

# Print and save
print(summary_short)
write.csv(summary_short, "./results/summary_stats_short_by_crop_model.csv", row.names = FALSE)
```

## Box-plots for each metric and crop/model
```{r}
# Pivot longer for metrics
df_long <- df_no_outliers %>%
  pivot_longer(cols = c(IoU, Precision, Recall, F1), names_to = "Metric", values_to = "Value")

# Plot
p_box <- ggplot(df_long, aes(x = crop, y = Value, fill = crop)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "black", outlier.color = "black") +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_fill_manual(values = c("wheat" = "#CEB888", "sorghum" = "#555960", "corn" = "#000000")) +
  labs(x = "Crop", y = "Value", title = "Boxplot of Metrics by Crop") +
  theme_minimal(base_size = 14) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), limits = c(0, 1), expand = c(0, 0.01)) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 13, face = "bold"),
    axis.title = element_text(size = 13, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

print(p_box)
ggsave("./results/boxplot_metrics_by_crop.png", plot = p_box, width = 7, height = 10, dpi = 600)
```

## Compare genotypes within each crop/model and get min/max mean per metric

```{r}
# Select the models
target_models <- tribble(
  ~crop,     ~modelname,
  "corn",    "SegFormer",
  "wheat",   "U-NET",
  "sorghum", "SegFormer"
)

# Make sure crop and modelname cases match your df_no_outliers
df_sel <- df_no_outliers %>%
  mutate(crop = trimws(tolower(as.character(crop))),
         modelname = trimws(as.character(modelname))) %>%
  # keep only selected pairs
  inner_join(target_models %>% mutate(crop = trimws(tolower(crop))),
             by = c("crop", "modelname"))

# Compute per-genotype mean, sd and CV (%) for each metric
df_genotype_mean <- df_sel %>%
  group_by(crop, modelname, genotype) %>%
  summarise(
    IoU_mean = mean(IoU, na.rm = TRUE),
    Precision_mean = mean(Precision, na.rm = TRUE),
    Recall_mean = mean(Recall, na.rm = TRUE),
    F1_mean = mean(F1, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Pivot CV columns and compute extremes per crop/model/metric
metrics_mean <- c("IoU_mean", "Precision_mean", "Recall_mean", "F1_mean")

df_extremes <- df_genotype_mean %>%
  pivot_longer(cols = all_of(metrics_mean), names_to = "metric", values_to = "Mean") %>%
  group_by(crop, modelname, metric) %>%
  summarise(
    min_mean = if (all(is.na(Mean))) NA_real_ else min(Mean, na.rm = TRUE),
    min_genotype = if (all(is.na(Mean))) NA else genotype[which.min(replace_na(Mean, Inf))],

    max_mean = if (all(is.na(Mean))) NA_real_ else max(Mean, na.rm = TRUE),
    max_genotype = if (all(is.na(Mean))) NA else genotype[which.max(replace_na(Mean, -Inf))],

    .groups = "drop"
  ) %>%
  # clean metric names for readability
  mutate(metric = case_when(
    metric == "IoU_mean" ~ "IoU Mean",
    metric == "Precision_mean" ~ "Precision Mean",
    metric == "Recall_mean" ~ "Recall Mean",
    metric == "F1_mean" ~ "F1 Mean",
    TRUE ~ metric
  )) %>%
  arrange(crop, modelname, metric)

# Round numerical CV columns for presentation
df_extremes <- df_extremes %>%
  mutate(
    min_mean = round(min_mean, 2),
    max_mean = round(max_mean, 2)
  )

# Print and save
print(df_extremes)
write.csv(df_extremes, "./results/genotype_mean_extremes_by_crop_model.csv", row.names = FALSE)
```

## Mean IoU by crop per collection date (grouped bar chart with error bars)

```{r}
# Select the models
target_models <- tribble(
  ~crop,     ~modelname,
  "corn",    "SegFormer",
  "wheat",   "U-NET",
  "sorghum", "SegFormer"
)

# Ensure collectiondate is a character and ordered 1..6
df_sel_dates <- df_no_outliers %>%
  mutate(crop = trimws(tolower(as.character(crop))),
         modelname = trimws(as.character(modelname))) %>%
  # keep only selected pairs
  inner_join(target_models %>% mutate(crop = trimws(tolower(crop))),
             by = c("crop", "modelname")) %>%
  mutate(collectiondate = as.character(collectiondate),
         collectiondate = trimws(collectiondate))

df_summary_date <- df_sel_dates %>%
  group_by(crop, collectiondate) %>%
  summarise(
    mean_IoU = mean(IoU, na.rm = TRUE),
    sd_IoU   = sd(IoU, na.rm = TRUE),
    n        = sum(!is.na(IoU)),
    lower    = pmax(0, mean_IoU - sd_IoU),
    upper    = pmin(1, mean_IoU + sd_IoU),
    .groups = "drop"
  )

# Make sure crop is a factor with correct levels for color mapping
df_summary_date$crop <- factor(df_summary_date$crop, levels = c("wheat", "sorghum", "corn"))

# Order collection dates including 6
df_summary_date$collectiondate <- factor(df_summary_date$collectiondate, levels = c("1","2","3","4","5","6"))

crops <- unique(as.character(df_summary_date$crop))

purdue_colors <- c(
  "black" = "#000000",
  "gold" = "#CEB888",
  "steel" = "#555960"
)

color_vals <- setNames(purdue_colors[seq_len(length(crops))], crops)

label_vals <- case_when(
  crops == "wheat" ~ "Wheat (U-NET)",
  crops == "sorghum" ~ "Sorghum (SegFormer)",
  crops == "corn" ~ "Corn (SegFormer)",
  TRUE                 ~ crops
)
names(label_vals) <- crops
# Dodge width
dodge_w <- 0.8

p_date_crop <- ggplot(df_summary_date, aes(x = collectiondate, y = mean_IoU, color = crop, group = crop)) +
  geom_line(size = 2, alpha = 0.9) +
  geom_point(size = 3.5, shape = 21, fill = "black", stroke = 1.5) +
  scale_color_manual(
    values = color_vals,
    labels = label_vals,
    breaks = names(label_vals),
    name = "Crop") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 20), limits = c(0, 1), expand = c(0, 0.01)) +
  labs(x = "Collection Date",
       y = "Mean IoU ± SD") +
  theme_minimal(base_size = 14) +
  theme(
    plot.background = element_rect(fill = purdue_colors["white"], color = NA),
    panel.background = element_rect(fill = purdue_colors["white"], color = NA),
    axis.text.x = element_text(color = purdue_colors["black"], size = 12),
    axis.text.y = element_text(color = purdue_colors["black"], size = 12),
    axis.title = element_text(color = purdue_colors["black"], size = 13, face = "bold"),
    plot.title = element_text(color = purdue_colors["black"], size = 16, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(color = purdue_colors["black"], size = 12, face = "bold"),
    legend.text = element_text(color = purdue_colors["black"], size = 11),
    panel.grid.major = element_line(color = purdue_colors["gold"], linewidth = 0.3),
    panel.grid.minor = element_blank()
  )

print(p_date_crop)
ggsave("./results/mean_IoU_by_crop_by_date.png", plot = p_date_crop, width = 8, height = 10, dpi = 600)
```

## Type I and III ANOVA (IoU) for selected crop-models

```{r}

df_sel <- df_no_outliers %>%
    filter(crop == trimws(tolower("corn")), modelname == "SegFormer")

df_sel <- df_sel %>%
    mutate(genotype = as.factor(genotype),
           collectiondate = as.factor(trimws(as.character(collectiondate))))

#fit <- tryCatch(aov(IoU ~ genotype * collectiondate, data = df_sel), error = function(e) e)
fit <- tryCatch(lm(IoU ~ genotype * collectiondate, data = df_sel), error = function(e) e)
if (inherits(fit, "error")) {
  message("ANOVA failed for corn - SegFormer: ", fit$message)
  next
}
out_prefix <- paste0("./results/anova_IoU_", "corn", "_", gsub("\\s+", "_", "SegFormer"))

#aov_summary <- summary(fit)[[1]]
aov_summary <- summary(fit)
aov_df <- as.data.frame(aov_summary)
#aov_df <- tibble::rownames_to_column(aov_df, var = "Term")
aov_df <- tibble::rownames_to_column(aov_df)
write.csv(aov_df, paste0(out_prefix, "_table.csv"), row.names = FALSE)

# Append human-readable output to a text log
sink(paste0(out_prefix, "_summary.txt"))
cat("ANOVA for", "corn", "-", "SegFormer", "\n\n")
print(summary(fit))
sink()
```

## t-tests

```{r}
# T-tests for min vs max genotypes across all performance metrics
ext_file <- "./results/genotype_mean_extremes_by_crop_model.csv"
if (!file.exists(ext_file)) stop("Extremes file not found: ", ext_file)
extremes <- read.csv(ext_file, stringsAsFactors = FALSE)

metrics_all <- c("IoU", "Precision", "Recall", "F1")
all_results <- list()

for (metric in metrics_all) {
  metric_label <- paste(metric, "Mean")
  ext_metric <- extremes %>%
    filter(metric == metric_label) %>%
    transmute(crop = trimws(tolower(as.character(crop))),
              genotype_min = as.character(min_genotype),
              genotype_max = as.character(max_genotype)) %>%
    distinct()

  metric_results <- list()
  for (i in seq_len(nrow(ext_metric))) {
    row <- ext_metric[i, ]
    crop_i <- row$crop
    g_min <- row$genotype_min
    g_max <- row$genotype_max
    out_prefix <- paste0("./results/ttest_", metric, "_", crop_i, "_", g_min, "_vs_", g_max)

    df_pair <- df_no_outliers %>%
      mutate(crop = trimws(tolower(as.character(crop))),
             genotype = as.character(genotype)) %>%
      filter(crop == crop_i, genotype %in% c(g_min, g_max)) %>%
      mutate(genotype = factor(genotype, levels = c(g_min, g_max)))

    n_min <- sum(df_pair$genotype == g_min, na.rm = TRUE)
    n_max <- sum(df_pair$genotype == g_max, na.rm = TRUE)

    if (n_min < 2 || n_max < 2) {
      metric_results[[i]] <- tibble(
        metric = metric, crop = crop_i, genotype_min = g_min, genotype_max = g_max,
        n_min = n_min, n_max = n_max, t_stat = NA_real_, df = NA_real_, p_value = NA_real_,
        mean_min = ifelse(n_min>0, mean(df_pair[[metric]][df_pair$genotype==g_min], na.rm=TRUE), NA_real_),
        mean_max = ifelse(n_max>0, mean(df_pair[[metric]][df_pair$genotype==g_max], na.rm=TRUE), NA_real_)
      )
      writeLines(paste("Not enough observations for t-test:", out_prefix), con = paste0(out_prefix, "_note.txt"))
      next
    }

    tt <- tryCatch(t.test(stats::as.formula(paste0(metric, " ~ genotype")), data = df_pair), error = function(e) e)
    if (inherits(tt, "error")) {
      metric_results[[i]] <- tibble(
        metric = metric, crop = crop_i, genotype_min = g_min, genotype_max = g_max,
        n_min = n_min, n_max = n_max, t_stat = NA_real_, df = NA_real_, p_value = NA_real_,
        mean_min = mean(df_pair[[metric]][df_pair$genotype==g_min], na.rm=TRUE),
        mean_max = mean(df_pair[[metric]][df_pair$genotype==g_max], na.rm=TRUE)
      )
      writeLines(paste("t.test error:", tt$message), con = paste0(out_prefix, "_error.txt"))
      next
    }

    res <- tibble(
      metric = metric,
      crop = crop_i,
      genotype_min = g_min,
      genotype_max = g_max,
      n_min = n_min,
      n_max = n_max,
      t_stat = as.numeric(tt$statistic),
      df = as.numeric(tt$parameter),
      p_value = as.numeric(tt$p.value),
      conf_low = as.numeric(tt$conf.int[1]),
      conf_high = as.numeric(tt$conf.int[2]),
      mean_min = mean(df_pair[[metric]][df_pair$genotype==g_min], na.rm=TRUE),
      mean_max = mean(df_pair[[metric]][df_pair$genotype==g_max], na.rm=TRUE)
    )

    metric_results[[i]] <- res

    sink(paste0(out_prefix, "_summary.txt"))
    cat("T-test for metric", metric, "comparing", g_min, "(min) vs", g_max, "(max)", "for crop", crop_i, "\n\n")
    print(tt)
    sink()
  }

  metric_tbl <- bind_rows(metric_results)
  write.csv(metric_tbl, paste0("./results/ttest_min_vs_max_", metric, "_by_crop.csv"), row.names = FALSE)
  all_results[[metric]] <- metric_tbl
}

tt_all <- bind_rows(all_results)
write.csv(tt_all, "./results/ttest_min_vs_max_by_crop_metric.csv", row.names = FALSE)
print(tt_all)
```

